---
title: "Aggregating Conveyal Regional Analyses"
author: "Emily Domanico, CTPS"
date: '2022-05-02'
output: github_document
---

# Aggregating Conveyal Regional Analyses

The purpose of this file is to work through an example of how to aggregate regional analysis results from Conveyal.
The example data used here is a regional analysis that measures access to non-emergency healthcare destinations in the MPO region.

## The goal 
The goal here is to work out how to aggregate results so that we can answer the question:
"How does access to basic healthcare services differ between EJ and non-EJ populations?"
 

```{r setup, warnings=F, message=FALSE, results='hide'}
# Raster processing
library(tidyverse)
library(sf)
library(stars)
library(mapview)
library(kableExtra)

# INPUTS ####
# read in census geog layers
mpo_census_tracts<- st_read("output/demographic_data.gpkg", layer= "tracts_acs_dec_2020") %>% 
  st_transform(3857)
boundary<-st_read("output/AggregationAreas.gpkg", layer= "MPO_Boundary") %>% 
  st_transform(3857) # needs to be psudo-mercator for raster operations
communitytypes<- st_read("output/AggregationAreas.gpkg", layer= "CommunityTypes") %>% st_transform(3857)
housingsubmarkets <- st_read("output/AggregationAreas.gpkg", layer="HousingSubmarkets") %>% st_transform(3857)

# read in geotiff
access_layer<- read_stars("data/ConveyalRuns/HealthCareTests/HealthCare_NonEmergency_AMPeak_TransitBusRT_45min_50pct.geotiff") 
destination <- "Non-emergency Healthcare"
time_period <- "AM Peak"
modes <- "Transit (Bus and Rapid Transit only)"
travel_time <- "45 minutes"

# read in destinations used for conveyal analysis
dest<- st_read("output/DestinationData.gpkg", layer= "healthcare_PT") %>% 
  st_transform(3857)

# note conveyal uses lodes workers home locations as a proxy for population density.
# the workers tiff is in stored in the same grid as the conveyal access output.
# To apply demographics we would still need to go through a process similar to option 2
# read in conveyal workers geotiff
# workers <- read_stars("data/")
```

First, crop the raster to the boundary extent.

```{r warnings=F, message=FALSE}
# CROP Access Raster to the MPO Boundary
access_layer_crop <- st_crop(access_layer, boundary)

```
```{r warnings=F, message=FALSE}
ggplot()+
  geom_stars(data=access_layer_crop, alpha = .7)+
  # geom_sf(data=access_contours, aes(color = value),size=.8, show.legend = F)+
  # geom_contour(data=access_layer_crop)+
  geom_sf(data=boundary,size=1,color="light gray", fill= NA)+
  coord_sf()+
  scale_fill_viridis_c(option = "D", na.value = "transparent",
                       name = paste0(destination, "\n Opportunities Accessible"))+
  # scale_color_viridis_c(option = "D", na.value = "transparent",
  #                      name = paste0(destination, "\n Opportunities Accessible"))+
  ggtitle(paste0("Access to ", destination, "\nwith ",
                                    modes, "\nin ", travel_time))+
  labs(caption= paste0("Time period: ", time_period))+
  theme_void()

```


## Option 1: Aggregate to the census geography
Find the average number of destinations accessible within a census geography. 
Then multiply average opportunities accessible from the tract by demographic population.

```{r warnings=F, message=FALSE}
# Option 1: find the average number of destinations accessible within a census tract

# This process finds the average number of opportunities accessible for cells that are within a given census tract geometry. Then multiplies the number of people from a given population in the tract by the average number of opportunities accessible to them.
# http://132.72.155.230:3838/r/combining-rasters-and-vector-layers.html#extracting-to-polygons-single-band

opps_by_tract<- access_layer_crop %>% 
  aggregate(mpo_census_tracts, mean, na.rm=T) %>% 
  st_as_sf() %>% 
  rename(opps= 1) # rename the avg opportunities column to "opps"

# join tract level demographic data by matching tract geog used to aggregate
opps_by_tract_demo<- mpo_census_tracts %>% 
  st_join(opps_by_tract, join = st_equals) %>% 
  mutate(
    people_opps = round(pop_dec*opps),
    opps_per_person = ifelse(pop_dec  == 0, 0, round(opps/pop_dec, 3)),
    min_opps = round(minority*opps),
    nonmin_opps = round(nonminority*opps),
    lowinc_opps = round(lowinc*opps),
    nonlowinc_opps = round(nonlowinc*opps))%>%
  mutate(id= row_number())
  #select(GEOID, ends_with("_opps"))
```


```{r warnings=F, message=FALSE}
opps_by_tract_demo %>% 
  select(GEOID, opps, pop_dec, ends_with("_opps")) %>% 
  st_drop_geometry() %>% 
  arrange(desc(min_opps)) %>% 
  head(10) %>% 
  kbl() %>% 
  kable_styling()
```
```{r warnings=F, message=FALSE}
mapview(select(opps_by_tract_demo,GEOID, NAME, ends_with("_opps")), zcol = "min_opps", layer.name = "Minority Opps")+
  mapview(select(opps_by_tract_demo,GEOID, NAME, ends_with("_opps")), zcol = "nonmin_opps", layer.name = "Nonminority Opps")+
  mapview(select(opps_by_tract_demo,GEOID, NAME, ends_with("_opps")), zcol = "lowinc_opps", layer.name = "Low-Income Opps")+
  mapview(select(opps_by_tract_demo,GEOID, NAME, ends_with("_opps")), zcol = "nonlowinc_opps", layer.name = "Non-low-income Opps")
```

To summarize region-wide:
```{r warnings=F, message=FALSE}

demo_summary_by_tract <- opps_by_tract_demo %>%
  st_drop_geometry() %>% 
  ungroup() %>% # or join and group by aggregation geometry instead
  summarize(people_opps = sum(people_opps, na.rm = T),
            min_opps = sum(min_opps, na.rm= T),
            nonmin_opps = sum(nonmin_opps, na.rm=T),
            lowinc_opps = sum(lowinc_opps, na.rm=T),
            nonlowinc_opps = sum(nonlowinc_opps, na.rm=T))

pct_opps_min_tract = demo_summary_by_tract$min_opps[1]/(demo_summary_by_tract$min_opps[1]+demo_summary_by_tract$nonmin_opps)
pct_opps_lowinc_tract = demo_summary_by_tract$lowinc_opps[1]/(demo_summary_by_tract$lowinc_opps[1]+demo_summary_by_tract$nonlowinc_opps)


demo_summary_by_tract_commtype <- opps_by_tract_demo %>% 
  st_join(communitytypes, largest = T) %>% # TODO: Fix munis that belong to mult subregions


```

```{r warnings=F, message=FALSE}
demo_summary_by_tract %>% 
  kbl() %>% 
  kable_styling()
```

Note some people-opps not assigned to demographics, because of ACS demo gaps and maybe treatment of region edges.
Overall, minority people have `r round(pct_opps_min_tract*100)`% of region-wide healthcare opportunities accessible by transit (Bus and RT).
And low-income people have ``r round(pct_opps_lowinc_tract*100)`% of region-wide healthcare opportunities accessible by transit (Bus and RT).


## Option 2: Pull demographics down to the raster cell

```{r warnings=F, message=FALSE}
# Option 2 allocate opportunities from each raster cell demo populations using areas overlap and 
tracts <- mpo_census_tracts %>% 
  mutate(area_tract = unclass(st_area(geom)))
# total_pop <- sum(tracts$pop_dec, na.rm=T)

access_crop_sf <- access_layer_crop %>% 
  st_as_sf() %>%  # convert raster to sf object (each row is a raster cell as vector)
  rename(opps = 1) %>% 
  mutate(area_cell = unclass(st_area(geometry)),
         cell_id = row_number())

#summary(access_crop_sf$opps)
# total_opps <- sum(access_crop_sf$opps)
# avg_opps <- mean(access_crop_sf$opps)

opps_by_cell_demo<- access_crop_sf %>% 
  st_intersection(select(tracts, 
                         GEOID, pop_dec, 
                         percent_min, percent_nonmin,
                         percent_lowinc, percent_nonlowinc,
                         area_tract)) %>% 
  mutate(area_int = unclass(st_area(geometry))) %>% 
  mutate(pct_overlap_tract = area_int/area_tract,
         pct_overlap_cell= area_int/area_cell) %>% 
  mutate(
    pop_part= pop_dec*pct_overlap_tract,
    opps_part = opps*pct_overlap_cell,
    pop_opps = pop_part*opps_part,
    min_opps = pop_opps*percent_min,
    nonmin_opps = pop_opps*percent_nonmin,
    lowinc_opps = pop_opps*percent_lowinc,
    nonlowinc_opps = pop_opps*percent_nonlowinc) %>% 
  # group back into whole cells
  group_by(cell_id) %>% 
  summarise(
    pop_opps = sum(pop_opps, na.rm=T),
    min_opps = sum(min_opps, na.rm=T),
    nonmin_opps = sum(nonmin_opps, na.rm=T),
    lowinc_opps = sum(lowinc_opps, na.rm=T),
    nonlowinc_opps = sum(nonlowinc_opps, na.rm=T),
    GEOIDs = paste0(unique(unlist(GEOID)), collapse = ' '),
    geometry= st_union(geometry)) %>% 
  st_as_sf()

```

```{r}
opps_by_cell_demo %>% 
  st_drop_geometry() %>% 
  arrange(desc(min_opps)) %>% 
  head() %>% 
  kbl() %>% 
  kable_styling()
```


To focus just on the cells associated with a specific tract:

```{r warnings=F, message=FALSE}
opps_by_cell_focus <- opps_by_cell_demo %>% 
  filter(grepl( "25025071101", GEOIDs))
```

```{r warnings=F, message=FALSE}
mapview(opps_by_cell_focus)
```

```{r}
demo_summary_by_cell <- opps_by_cell_demo %>% 
  st_drop_geometry() %>% 
  ungroup() %>% # or join and group by aggregation geometry instead
  summarize(people_opps = sum(pop_opps, na.rm = T),
            min_opps = sum(min_opps, na.rm= T),
            nonmin_opps = sum(nonmin_opps, na.rm=T),
            lowinc_opps = sum(lowinc_opps, na.rm=T),
            nonlowinc_opps = sum(nonlowinc_opps, na.rm=T))

pct_opps_min_cell = demo_summary_by_cell$min_opps[1]/(demo_summary_by_cell$min_opps[1]+demo_summary_by_cell$nonmin_opps)
pct_opps_lowinc_cell = demo_summary_by_cell$lowinc_opps[1]/(demo_summary_by_cell$lowinc_opps[1]+demo_summary_by_cell$nonlowinc_opps)
```
```{r}
demo_summary_by_cell %>% 
  kbl() %>% 
  kable_styling()
```

Note some people-opps not assigned to demographics, because of ACS demo gaps and maybe treatment of region edges.
Overall, minority people have `r round(pct_opps_min_cell*100)`% of region-wide healthcare opportunities accessible by transit (Bus and RT).
And low-income people have ``r round(pct_opps_lowinc_cell*100)`% of region-wide healthcare opportunities accessible by transit (Bus and RT).

Note: Percent of opportunities is the same for the two options, but the total people opps are not.


```{r warnings=F, message=FALSE, hide =T}
# test <- opps_by_cell_demo %>%
#   select(min_opps) %>%
#   st_as_stars()
# # 
# 
# access_contours <- st_contour(access_layer,
#                               na.rm=T,
#                               contour_lines = T,
#                               breaks = seq(10,30000, by= 1000) )
# colnames(access_contours)[1]<- "value"
# 
# ggplot()+
#   geom_stars(data=test, alpha = .7)+
#   # geom_sf(data=access_contours, aes(color = value),size=.8, show.legend = F)+
#   geom_contour(data=test)+
#   geom_sf(data=boundary,size=1,color="light gray", fill= NA)+
#   coord_sf()+
#   scale_fill_viridis_c(option = "D", na.value = "transparent",
#                        name = paste0(destination, "\n Opportunities Accessible"))+
#   # scale_color_viridis_c(option = "D", na.value = "transparent",
#   #                      name = paste0(destination, "\n Opportunities Accessible"))+
#   # ggtitle(paste0("Access to ", destination, "\nwith ",
#   #                                   modes, "\nin ", travel_time))+
#   # facet_wrap(~band)+
#   # coord_equal() +
#     # facet_wrap(~ba) +
#     theme_void() +
#     # scale_x_discrete(expand=c(0,0))+
#     # scale_y_discrete(expand=c(0,0))+
#   #
#   # labs(caption= paste0("Time period: ", time_period))+
#   theme_void()

```

